---
title: "BART Through Time Database Builder"
author: "David Ory"
output:
  html_document:
    theme: cosmo
    toc: yes
---

## Administration

#### Purpose
BART provides, on an annual basis, information on daily ridership.  An Excel macro is used to get the data out of Excel jail and into flat files.  This script combines the databases into a single file and creates a flat file of typical weekday statistics for use in Tableau. 

#### Outputs
1.  A condensed database of typical weekday transactions for use in Tableau (see ``Bart Through Time Typical Weekday.twb``).
2.  A larger database of transactions by date for general use.


## Procedure
#### Overhead

```{r overhead, results = 'hide'}
library(knitr)
library(stringr)
suppressMessages(library(dplyr))
library(timeDate)
library(chron)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Remote output file names
```{r file-names}
F_TYPICAL_R   = "D:/files/Box Sync/Share Data/bart-typical-weekday/BART Through Time Typical Weekday.Rdata"
F_TYPICAL_CSV = "D:/files/Box Sync/Share Data/bart-typical-weekday/BART Through Time Typical Weekday.csv"

F_ALL_R   = "D:/files/Box Sync/Share Data/bart-typical-weekday/BART Through Time.Rdata"
F_ALL_CSV = "D:/files/Box Sync/Share Data/bart-typical-weekday/BART Through Time.csv"

```


#### Methods
```{r methods}

Build_One_Month_Data <- function(input_month_string, input_year_string){
  
  file_path_string <- paste("M:/Data/Transit/", 
                            input_year_string, 
                            " Ridership/BART/by month/", 
                            input_year_string,
                            " ",
                            input_month_string,
                            ".csv",
                            sep = '')
  
  input_df <- read.table(file = file_path_string, header = TRUE, sep = ",", stringsAsFactors = FALSE)
  
  return(input_df)
  
}

Build_One_Year_Data <- function(input_year_string){
  
  
  month_string_array = c("April", "May", "September", "October", "November")
  
  # initialize with March data
  input.combine <- Build_One_Month_Data("March", input_year_string)
  
  for(month_string in month_string_array){
    working <- Build_One_Month_Data(month_string, input_year_string)
    input.combine <- rbind(input.combine, working)
  }
  
  
  input.combine <- mutate(input.combine, year = as.numeric(input_year_string))

}

```


### Data Reads
```{r all-years}

input.2015 <- Build_One_Year_Data('2015')
input.2014 <- Build_One_Year_Data('2014')
input.2013 <- Build_One_Year_Data('2013')
input.2012 <- Build_One_Year_Data('2012')
input.2011 <- Build_One_Year_Data('2011')
input.2010 <- Build_One_Year_Data('2010')
input.2005 <- Build_One_Year_Data('2005')
input.2000 <- Build_One_Year_Data('2000')

input.combine <- rbind(input.2000, input.2005, input.2010, 
                       input.2011, input.2012, input.2013, input.2014, input.2015)

```

### Combine and Clean up
```{r clean}

# Create date as date
combine.clean <- input.combine %>%
  mutate(date = as.Date(paste(day,month,year), format = "%d %B %Y"))

# Create holiday dummy (Labor Day, Memorial Day, Thanksgiving, and Veterans day are relevant) 
holiday_list  <- c("USLaborDay", "USMemorialDay", "USThanksgivingDay", "USVeteransDay")
holiday_dates <- dates(as.character(holiday(2000:2020, holiday_list)), format = "Y-M-D")

combine.clean <- combine.clean %>%
  mutate(holiday_dummy = is.holiday(date, holiday_dates))

# Create typical weekday dummy
combine.clean <- combine.clean %>%
  mutate(day_of_week = weekdays(date)) %>%
  mutate(typical_dummy = FALSE) %>%
  mutate(typical_dummy = ifelse(day_of_week == "Tuesday"   & !holiday_dummy, TRUE, typical_dummy)) %>%
  mutate(typical_dummy = ifelse(day_of_week == "Wednesday" & !holiday_dummy, TRUE, typical_dummy)) %>%
  mutate(typical_dummy = ifelse(day_of_week == "Thursday"  & !holiday_dummy, TRUE, typical_dummy))

# Give full station names
station.names = data.frame(abbrev = c("RM","EN","EP","NB","BK",
                                      "AS","MA","19","12","LM",
                                      "FV","CL","SL","BF","HY",
                                      "SH","UC","FM","CN","PH",
                                      "WC","LF","OR","RR","OW",
                                      "EM","MT","PL","CC","16",
                                      "24","GP","BP","DC","CM",
                                      "CV","ED","NC","WP","SS",
                                      "SB","SO","MB","WD","OA",
                                      "WS"), 
                           name = c("Richmond","El Cerrito del Norte","El Cerrito Plaza","North Berkeley","Berkeley",
                                    "Ashby","MacArthur","19th Street Oakland","12th Street Oakland","Lake Merritt",
                                    "Fruitvale","Coliseum","San Leandro","Bay Fair","Hayward",
                                    "South Hayward","Union City","Fremont","Concord","Pleasant Hill",
                                    "Walnut Creek","Lafayette","Orinda","Rockridge","West Oakland",
                                    "Embarcadero","Montgomery","Powell","Civic Center","16th Street Mission",
                                    "24th Street Mission","Glen Park","Balboa Park","Daly City","Colma",
                                    "Castro Valley","Dublin Pleasanton","North Concord","Pittsburg Bay Point","South City",
                                    "San Bruno","SF Airport","Millbrae","West Dublin", "Oakland Airport Connector",
                                    "Unknown")
                           )

# Join station names to origin station
station.orig <- station.names %>%
  select(orig_station = abbrev, orig_name = name)

combine.clean <- left_join(combine.clean, station.orig, by = 'orig_station', copy = FALSE)

station.dest <- station.names %>%
  select(dest_station = abbrev, dest_name = name)

combine.clean <- left_join(combine.clean, station.dest, by = 'dest_station', copy = FALSE)

# Remove unknown station for now (no riders, likely place holder for future station)
combine.clean <- combine.clean %>%
  filter(orig_name != "Unknown") %>%
  filter(dest_name != "Unknown")

```

### Summaries for typical weekdays
```{r summaries}

typical.clean <- combine.clean %>%
  filter(typical_dummy)

typical.sum <- typical.clean %>%
  group_by(orig_name, dest_name, year) %>%
  summarise(count_typical = n(), 
            sum_trips = sum(trips), 
            avg_trips = mean(trips), 
            min_trips = min(trips), 
            max_trips = max(trips), 
            stdev_trips = sd(trips), 
            median_trips = median(trips))

head(typical.sum, n = 25)

```

### Write to disk
```{r to disk}
# write typical file to disk for Tableau
save(typical.sum, file = F_TYPICAL_R)
write.csv(typical.sum, file = F_TYPICAL_CSV, row.names = FALSE, quote = F)

# write big file to disk
combine.write <- combine.clean %>%
  select(-day, -month, -year, -orig_station, -dest_station)

save(combine.write, file = F_ALL_R)
write.csv(combine.write, file = F_ALL_CSV, row.names = FALSE, quote = F)
```

