---
title: "BART Through Time Database Builder"
author: "David Ory"
output:
  html_document:
    theme: cosmo
    toc: yes
---

## Administration

#### Purpose
BART provides, on an annual basis, information on daily ridership.  I use an Excel macro to get the data out of Excel jail and into flat files.  This script combines the databases into as single file and creates a flat file of typical weekday statistics for use in Tableau. 

#### Outputs
1.  A condensed database for use in Tableau Public (see ``Bart Through Time Typical Weekday.twb``).
2.  A large database for SQL and Tableau(see ``BART Through Time.csv`` and ``[DavidOry].[BART Through Time]``)


## Procedure
#### Overhead

```{r overhead, results = 'hide'}
library(knitr)
library(stringr)
suppressMessages(library(dplyr))
library(timeDate)
library(chron)
```

```{r config, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

#### Remote output file names
```{r file-names}
F_TYPICAL_R   = "M:/Data/Transit/BART Through Time/BART Through Time Typical Weekday.Rdata"
F_TYPICAL_CSV = "M:/Data/Transit/BART Through Time/BART Through Time Typical Weekday.csv"

F_ALL_R   = "M:/Data/Transit/BART Through Time/BART Through Time.Rdata"
F_ALL_CSV = "M:/Data/Transit/BART Through Time/BART Through Time.csv"

```


#### Methods
```{r methods}

Build_One_Year_Data <- function(input_year_string){
  
  mar_string = paste("M:/Data/Transit/", input_year_string, " Ridership/BART/by month/", input_year_string, " March.csv", sep = '')
  apr_string = paste("M:/Data/Transit/", input_year_string, " Ridership/BART/by month/", input_year_string, " April.csv", sep = '')
  may_string = paste("M:/Data/Transit/", input_year_string, " Ridership/BART/by month/", input_year_string, " May.csv", sep = '')
  sep_string = paste("M:/Data/Transit/", input_year_string, " Ridership/BART/by month/", input_year_string, " September.csv", sep = '')
  oct_string = paste("M:/Data/Transit/", input_year_string, " Ridership/BART/by month/", input_year_string, " October.csv", sep = '')
  nov_string = paste("M:/Data/Transit/", input_year_string, " Ridership/BART/by month/", input_year_string, " November.csv", sep = '')
  
  input.mar <- read.table(file = mar_string, header = TRUE, sep = ",", stringsAsFactors = FALSE)
  input.apr <- read.table(file = apr_string, header = TRUE, sep = ",", stringsAsFactors = FALSE)
  input.may <- read.table(file = may_string, header = TRUE, sep = ",", stringsAsFactors = FALSE)
  input.sep <- read.table(file = sep_string, header = TRUE, sep = ",", stringsAsFactors = FALSE)
  input.oct <- read.table(file = oct_string, header = TRUE, sep = ",", stringsAsFactors = FALSE)
  input.nov <- read.table(file = nov_string, header = TRUE, sep = ",", stringsAsFactors = FALSE)
  
  input.combine <- rbind(input.mar, input.apr, input.may, input.sep, input.oct, input.nov)
  
  input.combine <- mutate(input.combine, year = as.numeric(input_year_string))

}

```


### Data Reads
```{r all-years}

input.2014 <- Build_One_Year_Data('2014')
input.2013 <- Build_One_Year_Data('2013')
input.2012 <- Build_One_Year_Data('2012')
input.2011 <- Build_One_Year_Data('2011')
input.2010 <- Build_One_Year_Data('2010')
input.2005 <- Build_One_Year_Data('2005')
input.2000 <- Build_One_Year_Data('2000')

input.combine <- rbind(input.2000, input.2005, input.2010, input.2011, input.2012, input.2013, input.2014)

```

### Combine and Clean up
```{r clean}

# Create date as date
combine.clean <- input.combine %>%
  mutate(date = as.Date(paste(day,month,year), format = "%d %B %Y"))

# Create holiday dummy (Labor Day, Memorial Day, Thanksgiving, and Veterans day are relevant) 
holiday_list  <- c("USLaborDay", "USMemorialDay", "USThanksgivingDay", "USVeteransDay")
holiday_dates <- dates(as.character(holiday(2000:2020, holiday_list)), format = "Y-M-D")

combine.clean <- combine.clean %>%
  mutate(holiday_dummy = is.holiday(date, holiday_dates))

# Create typical weekday dummy
combine.clean <- combine.clean %>%
  mutate(day_of_week = weekdays(date)) %>%
  mutate(typical_dummy = FALSE) %>%
  mutate(typical_dummy = ifelse(day_of_week == "Tuesday"   & !holiday_dummy, TRUE, typical_dummy)) %>%
  mutate(typical_dummy = ifelse(day_of_week == "Wednesday" & !holiday_dummy, TRUE, typical_dummy)) %>%
  mutate(typical_dummy = ifelse(day_of_week == "Thursday"  & !holiday_dummy, TRUE, typical_dummy))

# Give full station names
station.names = data.frame(abbrev = c("RM","EN","EP","NB","BK",
                                      "AS","MA","19","12","LM",
                                      "FV","CL","SL","BF","HY",
                                      "SH","UC","FM","CN","PH",
                                      "WC","LF","OR","RR","OW",
                                      "EM","MT","PL","CC","16",
                                      "24","GP","BP","DC","CM",
                                      "CV","ED","NC","WP","SS",
                                      "SB","SO","MB","WD"), 
                           name = c("Richmond","El Cerrito del Norte","El Cerrito Plaza","North Berkeley","Berkeley",
                                    "Ashby","MacArthur","19th Street Oakland","12th Street Oakland","Lake Merritt",
                                    "Fruitvale","Coliseum","San Leandro","Bay Fair","Hayward",
                                    "South Hayward","Union City","Fremont","Concord","Pleasant Hill",
                                    "Walnut Creek","Lafayette","Orinda","Rockridge","West Oakland",
                                    "Embarcadero","Montgomery","Powell","Civic Center","16th Street Mission",
                                    "24th Street Mission","Glen Park","Balboa Park","Daly City","Colma",
                                    "Castro Valley","Dublin Pleasanton","North Concord","Pittsburg Bay Point","South City",
                                    "San Bruno","SF Airport","Millbrae","West Dublin")
                           )

# Join station names to origin station
station.orig <- station.names %>%
  select(orig_station = abbrev, orig_name = name)

combine.clean <- left_join(combine.clean, station.orig, by = 'orig_station', copy = FALSE)

station.dest <- station.names %>%
  select(dest_station = abbrev, dest_name = name)

combine.clean <- left_join(combine.clean, station.dest, by = 'dest_station', copy = FALSE)

```

### Summaries for typical weekdays
```{r summaries}

typical.clean <- combine.clean %>%
  filter(typical_dummy)

typical.clean <- combine.clean[combine.clean$typical_dummy == TRUE, ]

typical.sum <- typical.clean %.%
  group_by(orig_name, dest_name, year) %.%
  summarise(count_typical = n(), 
            sum_trips = sum(trips), 
            avg_trips = mean(trips), 
            min_trips = min(trips), 
            max_trips = max(trips), 
            stdev_trips = sd(trips), 
            median_trips = median(trips))

head(typical.sum, n = 25)

```

### Write to disk
```{r to disk}
# write typical file to disk for Tableau
save(typical.sum, file = F_TYPICAL_R)
write.csv(typical.sum, file = F_TYPICAL_CSV, row.names = FALSE, quote = F)

# write big file to disk
combine.write <- combine.clean %>%
  select(-day, -month, -year, -orig_station, -dest_station)

save(combine.write, file = F_ALL_R)
write.csv(combine.write, file = F_ALL_CSV, row.names = FALSE, quote = F)


```

